\textit{Monami} as an oscillatory hydrodynamic instability in a submerged sea grass bed

### 2014-11-03 Paper/grass.tex
APS preprint style in 12 pts for uploading on the arXiv
Preprint format for revtex-4.1

\documentclass[aps,prl,twocolumn,superscriptaddress,sort&compress,10pt]{revtex4-1}  % for review and submission
% \documentclass[aps,preprint,10pt]{revtex4-1}  % for review and submission
%\documentclass[aps,twocolumn,floatfix,prl,10pt]{revtex4-1}
%\documentclass[letterpaper,10pt,prl,twocolumn,aps]{revtex4-1}
%\documentclass[aps,preprint,floatfix,prl]{revtex4-1}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{color}
\usepackage{longtable}
\usepackage{array}
\usepackage{amssymb}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}

@@

\begin{document}
\title{\textit{Monami} as an oscillatory hydrodynamic instability in a submerged sea grass bed}
\author{Ravi Singh}
\affiliation{Brown University, Providence RI 02912 USA}
% \author{L. Mahadevan}
% \affiliation{Harvard University, Cambridge MA 02138 USA}
\author{M. M. Bandi}
\affiliation{OIST Graduate University, Okinawa 904-0495, Japan}
\author{Amala Mahadevan}
\affiliation{Woods Hole Oceanographic Institution, Woods Hole MA 02543 USA}
\author{Shreyas Mandre}
\affiliation{Brown University, Providence RI 02912 USA}

@@

Although \monami ~is manifest in the motion of the grass, the drag exerted by the grass bed on the flow is central to the hypothesized instability. 
The instability and the resulting flow structures persist in lab experiments even when flexible grass mimics are replaced by rigid dowels \cite{Ghisal02,Nepf06}. 
Therefore, to develop the essential mathematical model, we assume the grass blades to be rigid, on average oriented vertically, and verify the assumption \textit{a posteriori}.
We show that the drag of the grass results in an instability mode different from the Kelvin-Helmholtz instability.
\begin{figure}
\includegraphics[scale=1]{Grass_Base_Nepf_shear}
% \includegraphics[width=15cm]{Grass_Base_Nepf_shear}
\caption{Schematic setup and comparison of our steady flow profile with that from the experiments by Ghisalberti and Nepf \cite{Nepf04} (Case-I from Table-1 with 1250 plants/m$^2$, 
plant height = 13.7$\pm 0.2$ cm and blade width of 0.64 cm)
 and its approximation with $U_0=7.28$ cm/s and $\delta = 5.02$ cm in our model. The grass extends up to $y=h_g$ in the water column of depth $2H$. 
The steady velocity profile can be decomposed into three regions, (i) A parabolic profile in the unvegetated region, (ii) a uniform profile deep within the vegetation, and (iii) a boundary layer of thickness $\delta$ near the grass top. 
The dependence of the boundary layer thickness (estimated as $|U/U_y|$ at $y=\hg$ from the numerical solution of \eqref{base_equ}) on the vegetation density parameter $\Rey \Ndg$ is shown in the inset.}
\label{basicflow}
\end{figure}
\begin{figure}
\begin{center}
\includegraphics[]{Critical_Re_vs_delta_noshear} \\
\vspace{-6mm} \hspace{-3mm}
\includegraphics[]{K_vs_shear_width_noshear}
% \includegraphics[width=12cm]{Critical_Re_vs_delta_noshear} \\
% \vspace{-6mm} \hspace{-5mm}
% \includegraphics[width=12cm]{K_vs_shear_width_noshear}
\end{center}
\caption{Critical Reynolds number (top) and the corresponding marginally stable wave number (bottom) for different submergence ratio as a function of vegetation density parametrized by the boundary layer thickness. 
Parameters estimated from experiments reported by Ghisalberti and Nepf \cite{Ghisal02} to exhibit or suppress synchronous waving are also included in the top panel. 
Inset compares experimental observations of the experimentally measured dominant frequency $f_o$ (in Hz) with the predictions $f_p$ from the solution of ~\eqref{Orr-somerfield}. 
The experimental data in the inset is obtained from publications by Ghisalberti and Nepf \cite{Ghisal02} and Vivoni \cite{Vivoni98}. 
In order to estimate the $\Rey$ for these experiments, a representative value of $\mu=0.1$ Pa~s was assumed.}
\label{Re_vs_delta}
\end{figure}

@@

\begin{table*}
\rowcolors{3}{tableShade}{white}  %% start alternating shades from 3rd row
\renewcommand{\arraystretch}{1.4}
 \begin{tabular}{l|c|c|c}
			& Kelvin-Helmholtz 				& Mode 1 		& Mode 2 \\ \hline
 Base velocity profile 	& $U(y) = U_0 \tanh(y/\delta)$			& \multicolumn{2}{c}{Equation \eqref{base_equ}} \\
 Domain 		& $-\infty < y < \infty$			& \multicolumn{2}{c}{$-1<y<1$} \\
 Inflection point	& exists at $y=0$				& \multicolumn{2}{c}{$U''(y)$ discontinuous at $y=\hg$} \\
 Shear layer thickness	& $\delta$					& \multicolumn{2}{c}{$\delta \sim  H\left(\Rey \Ndg \right)^{-1/3}$} \\
 Linearized dynamics	& $\left(\sigma+ikU\right) \left(D^2-k^2\right)\phi =  ikU_{yy}\phi$		& \multicolumn{2}{c}{Equation \eqref{Orr-somerfield}} \\
 Dense grass limit &  & Equation \eqref{eqn:mode1asymp} & Equation \eqref{eqn:mode2asymp}  \\
 Critical parameters	& none						& $\Rey \propto \Ndg^{1/2}$ 	& $\Rey \propto \Ndg$ \\
 Most unstable $k$ as $\delta \to 0$	& $\propto H/\delta$		& $\propto H/\delta$	& $O(1)$ \\
 Mode localized?	& yes, near $y=0$				& ~~yes, near $y=\hg$~~			& ~~no, spans water column~~
 \end{tabular}
 \caption{Comparison between Kelvin-Helmholtz instability and the two unstable modes resulting from solution of \ref{Orr-somerfield}.}
 \label{tab:comparison}
\end{table*}
To better understand the mechanism of waving and the origin of the threshold, we consider the behavior of the instability as a function of the vegetation drag characterized by $\Ndg$.
The fastest growing wavenumber first increases proportional to $H/\delta$, but at a critical vegetation density discontinuously jumps and remains $O(1)$ (see Fig.~\ref{Re_vs_delta}). 
To aid in explaining this behavior, we show heat maps of Re$(\sigma)$ as a function of $\Rey$ and $k$, for different $\hg/H$ and $\Ndg$ in Fig.~\ref{K_Re_sigma_set3}. 
The lowest $\Rey$ on the neutral curve (Re$(\sigma)=0$) sets the threshold. 
We observe that as $\Ndg$ increases, the unstable region splits into two; we refer to the unstable region with the higher $k$ as ``Mode 1'', and the one with the lower $k$ as ``Mode 2''. 
The unstable region for Mode 1, depending on $\hg/H$ either recedes to higher $\Rey$ or shrinks to zero size, as the vegetation density increases, causing the most unstable mode to transition discontinuously.

@@

On the other hand, the threshold condition for Mode 2 is numerically observed to be $\Rey \propto ({\delta}/{H})^{-3/2}$ (or $\Rey \propto \Ndg$) for $k\sim O(1)$, shown in Fig.~\ref{Re_vs_delta}, which can be understood by an asymptotic scaling analysis of ~\eqref{Orr-somerfield} by assuming $\Rey \gg 1$ but fixed $\Rey/\Ndg \sim O(1)$.
% Note that for large $\Ndg$, the non-dimensional steady flow velocity inside the vegetation is $U_g \sim (\Rey \Ndg)^{-1/2}$. 
In this limit ~\eqref{Orr-somerfield} simplifies to 
\begin{align}
\begin{split}
\sigma\left(D^2-k^2\right)\phi = -2{(\Ndg/\Rey)^{1/2}}D^2\phi,  \quad &\text{ for $y<\hg$}  \\
\left(\sigma+ikU\right) \left(D^2-k^2\right)\phi =  ikU_{yy}\phi, \quad &\text{ for $y>\hg$}.
\end{split}
\label{eqn:mode2asymp}
\end{align}
\begin{figure}
\includegraphics[]{Asymptotic_noshear}
\caption{Plot of mode shape $|\phi|$ for different shear layer thickness in limit of small shear layer thickness for representative values of vegetation density. 
Mode 1 is shown in solid and Mode 2 is shown in dashed. The parameters Mode 2 shapes are chosen such that $\Rey \gg 1$, $\Ndg \gg 1$ (specified in terms of $\delta/H$) but $\Rey/\Ndg = O(1)$. 
The mode shapes approach each other for these small values of $\delta/H$ indicating that the large vegetation density asymptote is reached. Mode 1 shapes appear self-similar in shape as $\delta\to 0$, 
and are compared to each other on a rescaled coordinate in the inset. 
Inset shows $|\phi|$ for Mode 1 as a function of $(y-\hg)/\delta$. The modes approach a universal shape, indicating that an asymptotic limit has been reached. 
The limit is not yet reached for the case $\delta/H = 0.005$ due to the influence of bottom boundary; the vegetation height in this case is comparable to the boundary layer thickness.}
\label{Asymptotic_mode}
\end{figure}
Since $\Rey/\Ndg$ is the only remaining parameter, the mode shape converges in the aforementioned limit, in agreement with our numerical results for Mode 2 shown in Fig.~\ref{Asymptotic_mode}. 
We interpret Mode 2 as the instability of an inviscid flow, with the vegetation modeled by a continuum drag field, and for which the boundary layer near the top of the vegetation plays no role. The only remaining parameter $\Rey/\Ndg$ 
sets the threshold, leading to the asymptotic behavior $\Rey \propto \Ndg$ (or $\Rey \sim ({\delta}/{H})^{-3/2}$).
%On the other hand, Mode 1 asymptotically localizes to the boundary layer near the grass tip, and exhibits a different asymptotic behavior with $k \sim O(H/\delta)$, and $\Rey \sim (H/\delta)$ (or $\Rey \propto {\Ndg}^{1/2}$) at the threshold.

>>>>

% \documentclass[aps,prl,twocolumn,superscriptaddress,sort&compress,10pt]{revtex4-1}  % for review and submission
\documentclass[aps,preprint,12pt]{revtex4-1}  % for review and submission
%\documentclass[aps,twocolumn,floatfix,prl,10pt]{revtex4-1}
%\documentclass[letterpaper,10pt,prl,twocolumn,aps]{revtex4-1}
%\documentclass[aps,preprint,floatfix,prl]{revtex4-1}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{color}
\usepackage{longtable}
\usepackage{array}
\usepackage{amssymb}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}

@@

\begin{document}
\title{\textit{Monami} as an oscillatory hydrodynamic instability in a submerged sea grass bed}
\author{Ravi Singh}t
\affiliation{Brown University, Providence RI 02912 USA}
% \author{L. Mahadevan}
% \affiliation{Harvard University, Cambridge MA 02138 USA}
\author{M. M. Bandi}
\affiliation{OIST Graduate University, Okinawa 904-0495, Japan}
\author{Amala Mahadevan}
\affiliation{Woods Hole Oceanographic Institution, Woods Hole MA 02543 USA}
\author{Shreyas Mandre}
\affiliation{Brown University, Providence RI 02912 USA}

@@

Although \monami ~is manifest in the motion of the grass, the drag exerted by the grass bed on the flow is central to the hypothesized instability. 
The instability and the resulting flow structures persist in lab experiments even when flexible grass mimics are replaced by rigid dowels \cite{Ghisal02,Nepf06}. 
Therefore, to develop the essential mathematical model, we assume the grass blades to be rigid, on average oriented vertically, and verify the assumption \textit{a posteriori}.
We show that the drag of the grass results in an instability mode different from the Kelvin-Helmholtz instability.
\begin{figure}
% \includegraphics[scale=1]{Grass_Base_Nepf_shear}
\includegraphics[width=15cm]{Grass_Base_Nepf_shear}
\caption{Schematic setup and comparison of our steady flow profile with that from the experiments by Ghisalberti and Nepf \cite{Nepf04} (Case-I from Table-1 with 1250 plants/m$^2$, 
plant height = 13.7$\pm 0.2$ cm and blade width of 0.64 cm)
 and its approximation with $U_0=7.28$ cm/s and $\delta = 5.02$ cm in our model. The grass extends up to $y=h_g$ in the water column of depth $2H$. 
The steady velocity profile can be decomposed into three regions, (i) A parabolic profile in the unvegetated region, (ii) a uniform profile deep within the vegetation, and (iii) a boundary layer of thickness $\delta$ near the grass top. 
The dependence of the boundary layer thickness (estimated as $|U/U_y|$ at $y=\hg$ from the numerical solution of \eqref{base_equ}) on the vegetation density parameter $\Rey \Ndg$ is shown in the inset.}
\label{basicflow}
\end{figure}
\begin{figure}
\begin{center}
% \includegraphics[]{Critical_Re_vs_delta_noshear} \\
% \vspace{-6mm} \hspace{-3mm}
% \includegraphics[]{K_vs_shear_width_noshear}
\includegraphics[width=12cm]{Critical_Re_vs_delta_noshear} \\
\vspace{-6mm} \hspace{-5mm}
\includegraphics[width=12cm]{K_vs_shear_width_noshear}
\end{center}
\caption{Critical Reynolds number (top) and the corresponding marginally stable wave number (bottom) for different submergence ratio as a function of vegetation density parametrized by the boundary layer thickness. 
Parameters estimated from experiments reported by Ghisalberti and Nepf \cite{Ghisal02} to exhibit or suppress synchronous waving are also included in the top panel. 
Inset compares experimental observations of the experimentally measured dominant frequency $f_o$ (in Hz) with the predictions $f_p$ from the solution of ~\eqref{Orr-somerfield}. 
The experimental data in the inset is obtained from publications by Ghisalberti and Nepf \cite{Ghisal02} and Vivoni \cite{Vivoni98}. 
In order to estimate the $\Rey$ for these experiments, a representative value of $\mu=0.1$ Pa~s was assumed.}
\label{Re_vs_delta}
\end{figure}

@@

\begin{table*}
\footnotesize
\rowcolors{3}{tableShade}{white}  %% start alternating shades from 3rd row
\renewcommand{\arraystretch}{1.2}
 \begin{tabular}{l|c|c|c}
			& Kelvin-Helmholtz 				& Mode 1 		& Mode 2 \\ \hline
 Base velocity profile 	& $U(y) = U_0 \tanh(y/\delta)$			& \multicolumn{2}{c}{Equation \eqref{base_equ}} \\
 Domain 		& $-\infty < y < \infty$			& \multicolumn{2}{c}{$-1<y<1$} \\
 Inflection point	& exists at $y=0$				& \multicolumn{2}{c}{$U''(y)$ discontinuous at $y=\hg$} \\
 Shear layer thickness	& $\delta$					& \multicolumn{2}{c}{$\delta \sim  H\left(\Rey \Ndg \right)^{-1/3}$} \\
 Linearized dynamics	& $\left(\sigma+ikU\right) \left(D^2-k^2\right)\phi =  ikU_{yy}\phi$		& \multicolumn{2}{c}{Equation \eqref{Orr-somerfield}} \\
 Dense grass limit &  & Equation \eqref{eqn:mode1asymp} & Equation \eqref{eqn:mode2asymp}  \\
 Critical parameters	& none						& $\Rey \propto \Ndg^{1/2}$ 	& $\Rey \propto \Ndg$ \\
 Most unstable $k$ as $\delta \to 0$	& $\propto H/\delta$		& $\propto H/\delta$	& $O(1)$ \\
 Mode localized?	& yes, near $y=0$				& yes, near $y=\hg$			& no, spans water column
 \end{tabular}
 \caption{Comparison between Kelvin-Helmholtz instability and the two unstable modes resulting from solution of \ref{Orr-somerfield}.}
 \label{tab:comparison}
\end{table*}
To better understand the mechanism of waving and the origin of the threshold, we consider the behavior of the instability as a function of the vegetation drag characterized by $\Ndg$.
The fastest growing wavenumber first increases proportional to $H/\delta$, but at a critical vegetation density discontinuously jumps and remains $O(1)$ (see Fig.~\ref{Re_vs_delta}). 
To aid in explaining this behavior, we show heat maps of Re$(\sigma)$ as a function of $\Rey$ and $k$, for different $\hg/H$ and $\Ndg$ in Fig.~\ref{K_Re_sigma_set3}. 
The lowest $\Rey$ on the neutral curve (Re$(\sigma)=0$) sets the threshold. 
We observe that as $\Ndg$ increases, the unstable region splits into two; we refer to the unstable region with the higher $k$ as ``Mode 1'', and the one with the lower $k$ as ``Mode 2''. 
The unstable region for Mode 1, depending on $\hg/H$ either recedes to higher $\Rey$ or shrinks to zero size, as the vegetation density increases, causing the most unstable mode to transition discontinuously.

@@

On the other hand, the threshold condition for Mode 2 is numerically observed to be $\Rey \propto ({\delta}/{H})^{-3/2}$ (or $\Rey \propto \Ndg$) for $k\sim O(1)$, shown in Fig.~\ref{Re_vs_delta}, which can be understood by an asymptotic scaling analysis of ~\eqref{Orr-somerfield} by assuming $\Rey \gg 1$ but fixed $\Rey/\Ndg \sim O(1)$.
% Note that for large $\Ndg$, the non-dimensional steady flow velocity inside the vegetation is $U_g \sim (\Rey \Ndg)^{-1/2}$. 
In this limit ~\eqref{Orr-somerfield} simplifies to 
\begin{align}
\begin{split}
\sigma\left(D^2-k^2\right)\phi = -2{(\Ndg/\Rey)^{1/2}}D^2\phi,  \quad &\text{ for $y<\hg$}  \\
\left(\sigma+ikU\right) \left(D^2-k^2\right)\phi =  ikU_{yy}\phi, \quad &\text{ for $y>\hg$}.
\end{split}
\label{eqn:mode2asymp}
\end{align}
\begin{figure}
% \includegraphics[]{Asymptotic_noshear}
\includegraphics[width=12cm]{Asymptotic_noshear}
\caption{Plot of mode shape $|\phi|$ for different shear layer thickness in limit of small shear layer thickness for representative values of vegetation density. 
Mode 1 is shown in solid and Mode 2 is shown in dashed. The parameters Mode 2 shapes are chosen such that $\Rey \gg 1$, $\Ndg \gg 1$ (specified in terms of $\delta/H$) but $\Rey/\Ndg = O(1)$. 
The mode shapes approach each other for these small values of $\delta/H$ indicating that the large vegetation density asymptote is reached. Mode 1 shapes appear self-similar in shape as $\delta\to 0$, 
and are compared to each other on a rescaled coordinate in the inset. 
Inset shows $|\phi|$ for Mode 1 as a function of $(y-\hg)/\delta$. The modes approach a universal shape, indicating that an asymptotic limit has been reached. 
The limit is not yet reached for the case $\delta/H = 0.005$ due to the influence of bottom boundary; the vegetation height in this case is comparable to the boundary layer thickness.}
\label{Asymptotic_mode}
\end{figure}
Since $\Rey/\Ndg$ is the only remaining parameter, the mode shape converges in the aforementioned limit, in agreement with our numerical results for Mode 2 shown in Fig.~\ref{Asymptotic_mode}. 
We interpret Mode 2 as the instability of an inviscid flow, with the vegetation modeled by a continuum drag field, and for which the boundary layer near the top of the vegetation plays no role. The only remaining parameter $\Rey/\Ndg$ 
sets the threshold, leading to the asymptotic behavior $\Rey \propto \Ndg$ (or $\Rey \sim ({\delta}/{H})^{-3/2}$).
%On the other hand, Mode 1 asymptotically localizes to the boundary layer near the grass tip, and exhibits a different asymptotic behavior with $k \sim O(H/\delta)$, and $\Rey \sim (H/\delta)$ (or $\Rey \propto {\Ndg}^{1/2}$) at the threshold.
